---
title: "Assignment 3A - Global Baseline Estimate"
author: "Ellie Wilser"
editor: visual
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    code-summary: "Show code"
    code-tools: true
    df-print: paged
    smooth-scroll: true
    embed-resources: true
---

## Introduction

For this assignment I will use the data in my previous assignment that I collected from my friends. I copied the code that I used to load the data from PostgreSQL, and will implement a Global Baseline Estimate algorithm. This is a non-personalized recommendation system.

Using the provided document ([Link to Document](https://brightspace.cuny.edu/content/enforced/1344367-SPS01_DATA_607_1269_1_34036/MovieRatings.xlsx?ou=1344367)), I will create a algorithm that predicts the missing values in my data. Missing data points are from where users did not see the movie. This model will be built upon in further assignments by using personalized recommendations systems.

## Approach

I will start this assignment by loading in my data from PostgreSQL and putting the data into a Tidy format. After, the data will be reshaped to the necessary format for the algorithm.

To calculate the recommendations, I will first calculate the global average, user averages, and movies averages. Using these values I will predict missing user rating by finding the user and movie biases, then adding to the global mean.

The only challenges I anticipate is reformatting the data, and the small number of friends surveyed. Since I only surveyed a limited about of people, the biases I calculate may not be the best representation to judge against the global mean.

## Code

### Loading Data (Copied from Assignment 2)

```{r Libraries, message=FALSE, warning=FALSE}
library(DBI)
library(RPostgres)
library(tidyverse)
library(readr)
library(dplyr)

```

```{r Connecting To DB}
con <- dbConnect(
  RPostgres::Postgres(),
  dbname   = "Assignment_2",
  host     = "localhost",
  port     = 5432,
  user     = "postgres"
)
```

```{r Loading Data}
ratings <- dbReadTable(con, "ratings")
ratings
ratings_tidy <- ratings %>%
  pivot_longer(
    cols = ends_with("_rating"),
    names_to = "movie",
    values_to = "rating",
    values_drop_na = FALSE
  ) %>%
  mutate(movie = gsub("_rating$", "", movie))

ratings_tidy
```

## Global Baseline Estimate Implementation

To implement the global baseline estimate algorithm, I first created dataframes for user averages and movies averages, and the global average consent. These variables will be used in algorithm.

```{r User and Movie Averages}

global_average <- mean(ratings_tidy$rating, na.rm=TRUE)
user_average <- ratings_tidy %>%
  group_by(friend_name) %>%
  summarize(avg_rating_user = mean(rating, na.rm=TRUE))
movie_average <- ratings_tidy %>%
  group_by(movie) %>%
  summarize(avg_rating_movie = mean(rating, na.rm=TRUE))
```

Since the algorithm will be repeated each time a prediction will be made, I decided to create a function. Each time the function is called a prediction will be calculated for the given user and movie.

```{r Define Function}

get_baseline_rating <- function(user,mov){
  u_avg <- user_average %>% filter(friend_name == user) %>% pull(avg_rating_user)
  m_avg <- movie_average %>% filter(movie == mov) %>% pull(avg_rating_movie)
  
  prediction <- global_average + (u_avg-global_average) + (m_avg-global_average)
  
  return(round(prediction,2))
  
}
```

## Predicting Ratings and Recommendations

Donna, and Alice are friends that are trying to decide what movie to see next. I will go through both and tell them which movie I would recommend for them.

### Alice's Recommendation

Using the algorithm, Alice's predicted rating for Spider Man was 1.95 and Obsession was 2.5. This means I would recommend Alice see Obsession next instead of Spider Man.

```{r Predicting Ratings - Alice}
get_baseline_rating("Alice", "spider_man")
get_baseline_rating("Alice", "obession")
```

### Donna's Recommendation

The baseline rating for Donna was a tie between weapons and mayday, meaning that either movie would be a good recommendation for her.

```{r Predicting Ratings - Donna}
get_baseline_rating("Donna", "weapons")
get_baseline_rating("Donna", "mayday")
```

## Conclusion

This algorithm does a great job providing a baseline for future recommendations. The algorithm relies heavily on the movie's average for all users. If the average for the one movie is the same as another, the algorithm will give the same predicted rating for the same user. This is because when making a recommendation for a user to decided between 2 different movies, the only variable that changes is the movie average. Personalized recommendations would give a better recommendation for users trying to decide what to watch next.
