---
title: "Data 607: Assignment 1"
author: "Ellie Wilser"
format: html
execute:
  echo: true
---

## Approach

For Assignment 1, I plan to use a data source that is the CO~2~ emissions by country from 1750 to current. More information about the data source can be found in the *Source* section. The data source contains multiple columns with data ranging from country demographic data to energy consumption. I plan to explore the columns relating to temperature change. I will remove the other columns, modify columns when necessary, and create a few graphs to explore the data. A data challenge I might run into is when creating the graphs. I plan to familiar my self with the ggplot2 library using the R for Data Science textbook.

## Introduction

The source for the assignment is the article *CO~2~ Emissions: How much CO~2~ does the world emit? Which countries emit the most.* Hannah Ritchie and Max Roser evaluate which countries are the largest producers of CO2 by using various data visualizations. The numerous visuals look at raw emissions by year, normalized emissions using population totals, and cumulative sums going back to the 1800s. The authors share the data for others to analyze.

The following link connects to the Github Repo where the data is being stored: <https://raw.githubusercontent.com/EllieWilser/Data_607/refs/heads/main/Assignment_1/owid-co2-data.csv>






## Code

### Library
``` {r}
library(readr)
library(dplyr)
library(tidyverse)

```

### Load Data from Github

``` {r}

url <- "https://raw.githubusercontent.com/EllieWilser/Data_607/refs/heads/main/Assignment_1/owid-co2-data.csv"

co2_emissions <- read_csv(
  file = url,
  show_col_types = FALSE,
  progress = FALSE
)
```

### Data Cleaning

In order to analyze the data file, a few cleaning tasks had to be done. First, I used dplyr to select the columns that I needed, then I removed rows which had NA values in them. This worked with the structure of the data file, and no needed data was lost. Since the temperature change from ghg column was cumulative, I removed all years besides 2022 and then removed the "World" entry to avoid double counting.

```{r}
co2_emissions <- select(co2_emissions, "country", "year","population","gdp","temperature_change_from_ghg")
co2_emissions_clean <- co2_emissions[rowSums(is.na(co2_emissions)) == 0,]
co2_emissions_clean <- filter(co2_emissions_clean, year == 2022)
co2_emissions_clean$temperature_change_from_ghg = as.numeric(co2_emissions_clean$temperature_change_from_ghg)
co2_emissions_clean <- filter(co2_emissions_clean, country != "World")

co2_emissions_clean
```

### Basic Exploratory Data Analysis

To better understand the data, I calculated the total, mean and median of the temperature change from ghg column.

```{r}

sum(co2_emissions_clean$temperature_change_from_ghg)
mean(co2_emissions_clean$temperature_change_from_ghg,na.rm = FALSE)
median(co2_emissions_clean$temperature_change_from_ghg,na.rm = FALSE)
```

### Graphs

Using ggplot, I created a pie chart of the temperature change from ghg by country. As a reference, I used r graph gallery which was a great resource for learning how to create a pie chart in ggplot.

The link to r graph gallery: <https://r-graph-gallery.com/piechart-ggplot2.html>

```{r}
plotting_data <- co2_emissions_clean %>%
  mutate(rank = rank(-temperature_change_from_ghg),
         country = ifelse(rank <= 10,country,'Other')) %>%
  group_by(country) %>%
  summarise(temperature_change_from_ghg = sum(temperature_change_from_ghg))

ggplot(plotting_data, aes(x="", y=temperature_change_from_ghg, fill=country)) +
  geom_bar(stat="identity",width=1) +
  coord_polar("y", start=0) +
  theme_void() +
  geom_text(aes(x =1.6, label = paste0(temperature_change_from_ghg,"°C")), position = position_stack(vjust = .5))+
  labs(
    title = "Temperature Change from GHG by Country"
  )


```
## Citation

**Holtz, Yan. (2025)** *ggplot2 Piechart*. The R Graph Gallery. \
<https://r-graph-gallery.com/piechart-ggplot2.html>

**Ritchie, Hannah, and Max Roser. (2020).** *CO₂ emissions*. Our World in Data.\
<https://archive.ourworldindata.org/20260810-163104/co2-emissions.html>

