---
title: "Data 607: Assignment 2A"
author: "Ellie Wilser"
editor: visual
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    code-summary: "Show code"
    code-tools: true
    df-print: paged
    smooth-scroll: true
    embed-resources: true
---

## Approach

For this assignment, I plan to create a survey with Google Forms that asks people to rate 6 different movies. The choices will be 1 through 5 and another option with "Did not see". When handling "Did not see", I will store that entry as an NA or NULL in my data table. When performing analysis, I will leave out that rating so not to have an effect.

I plan to load the data from the survey into SQL and then connect the SQL server to my R environment to query the database directly. I anticipate a slight challenge getting this connection set up, so I might have to export the data from the database and load it into R.

After the data is in my R environment, I will compute average ratings and count ratings per user and movie.

## Code

```{r Libraries}
library(DBI)
library(RPostgres)
library(tidyverse)
library(readr)
library(dplyr)
library(tidyverse)

```

### Connecting to Postgres

In my approach section, I was wary if I would be able to connect to PostgreSQL in my R environment. After researching, I was able to successfully connect RStudio to my database.

```{r Connecting To DB}
con <- dbConnect(
  RPostgres::Postgres(),
  dbname   = "Assignment_2",
  host     = "localhost",          # Use IP or server address if remote
  port     = 5432,                 # 5432 is the default PostgreSQL port
  user     = "postgres"
)
```

### Loading and Transforming to Tidy Formatting

When I ingested my data into PostgreSQL, I formatted in the way of how the responses came in. Each row represented a survey, but I switched to long form in order for the data to fit Tidy formatting. Now each row represents a single rating. See below the table before and after cleaning:

```{r Loading Data and Transforming}

ratings <- dbReadTable(con, "ratings")
ratings
ratings_tidy <- ratings %>%
  pivot_longer(
    cols = ends_with("_rating"),
    names_to = "movie",
    values_to = "rating",
    values_drop_na = FALSE
  ) %>%
  mutate(movie = gsub("_rating$", "", movie))

ratings_tidy
```

### Summarizing Data

When summarizing the data I filtered out the NA values in order to not cause an error. The user_summary table has total number of ratings and average rating by user, and movie summary has the same but for the movies. I included the .groups = "drop" to avoid a warning message and not cause future errors when interacting with the data frame.

```{r User and Move Summary}

user_summary <- ratings_tidy %>%
  filter(!is.na(rating)) %>%
  group_by(friend_id, friend_name) %>%
  summarise(
    total_ratings = n(),
    avg_rating    = mean(rating),
    .groups       = "drop"
  )

movie_summary <- ratings_tidy %>%
  filter(!is.na(rating)) %>%
  group_by(movie) %>%
  summarise(
    total_ratings = n(),
    avg_rating    = mean(rating),
    .groups       = "drop"
  )

user_summary

movie_summary

```
