---
title: "Data 607: Assignment 2B"
author: "Ellie Wilser"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    df-print: paged
    smooth-scroll: true
    embed-resources: true
editor: visual
---

## Approach

For this assignment, I plan to go through the planned out code chucks below which matches the assignment tasks. Before starting to code, I plan to read through the provided PDF and get a refresh on binary classification. I have worked with these model's before, but the PDF will be useful. I don't think I will have too many difficulties with this assignment, but I want to challenge myself to use Git in terminal and make a graph using ggplot 2.

## Code

### Loading Data and Libraries

```{r}
library(readr)
library(dplyr)
library(tidyverse)
```

```{r}
url <- "https://raw.githubusercontent.com/EllieWilser/Data_607/refs/heads/main/Assignment_2/penguin_predictions.csv"

penguin_pred <- read_csv(
  file = url,
  show_col_types = FALSE,
  progress = FALSE
)
  
```

### Data Exploration

```{r}

```

### Graph

```{r}

```

### Null Error Rate

```{r}

```

### Confusion Matrices at Multiple Thresholds

```{r}

```

### Performance Metrics

```{r}

```

### **Threshold Use Cases**

```{r}

```
